---
title: "Modeling using Logistic Regression"
format: html
editor: visual
---

```{r}
library(tidyverse)
library(tidymodels)
library(vip)
library(readr)
library(here)
library(themis)
library(yardstick)
library(ggplot2)
library(dplyr)
```


#Additional model preperation


We take *ufo_model_data* which was the main dataframe, and we join the "places" dataframe to it.
```{r}
ufo_final_model_data <- ufo_model_data %>%
  left_join(places_model_data, by = c("city", "state", "country_code"))

ufo_final_model_data
```


We have selected only several columns that we determined to be the most necessary in our predictions: shape, duration (in seconds), part of day, population of the city it took place and the geographical data.
```{r}
ufo_final_model_data <- ufo_final_model_data %>%
  select(shape, duration_seconds, day_part, population, elevation_m, latitude, longitude) %>%
  filter(!is.na(shape)) %>%
  mutate(shape = factor(shape))

ufo_final_model_data
```


We wanted our model to predict the type of shape a UFO sighting could have. This dataset had a lot of different shapes, so we decided to only use the 4 most common shapes
```{r}
top_shapes <- ufo_final_model_data %>%
  count(shape, sort = TRUE) %>%
  slice(1:4) %>%
  pull(shape)

top_shapes
```

Filtering the dataframe based on the 4 most common shapes
```{r}
ufo_final_model_data <- ufo_final_model_data %>%
  filter(shape %in% top_shapes) %>%
  mutate(shape = factor(shape))

ufo_final_model_data
```


Five fold cross validation set-up
```{r}
set.seed(123)
split <- initial_split(ufo_final_model_data, strata = day_part)
train_data <- training(split)
test_data <- testing(split)

```


Recipe
```{r}
rec <- recipe(shape ~ duration_seconds + day_part + population + elevation_m + latitude + longitude,
              data = ufo_final_model_data) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors()) %>%
  step_normalize(all_numeric_predictors())

rec
```

#Random Forest

Our second model type is a random forest. We tried it as it also fit our case and would be (in theory) more advanced than multinomal regression.

Model
```{r}
rf_model <- rand_forest(mode = "classification", trees = 500) %>%
  set_engine("ranger")

rf_model
```

Workflow
```{r}
wf_rf <- workflow() %>%
  add_recipe(rec) %>%
  add_model(rf_model)
```


Model fitting
```{r}
rf_fit <- fit(wf_rf, data = train_data)
```


Predictions
```{r}
rf_preds <- predict(rf_fit, test_data) %>%
  bind_cols(test_data) %>%
  mutate(
    shape = factor(shape),
    .pred_class = factor(.pred_class, levels = levels(shape))
  )
```


Trained model metrics. Generally pretty poor results
```{r}
metrics <- metric_set(accuracy, precision, recall, f_meas)

metrics(rf_preds, truth = shape, estimate = .pred_class, estimator = "macro_weighted")
```



Confusion matrix - it clearly shows that the model does not want to predict anything other than *circle* and *light*. This time however there is definetly an improvement.
```{r}
conf_mat(rf_preds, truth = shape, estimate = .pred_class)
```



