---
title: "Modeling using Logistic Regression"
format: html
editor: visual
---

## Preparing the data

```{r}
library(tidyverse)
library(tidymodels)
library(vip)
library(readr)
library(here)
<<<<<<< HEAD
library(themis)
=======
>>>>>>> cb33d2047023d675e060afd2b401560b5c0ebe59
library(yardstick)
library(ggplot2)
library(dplyr)
```


#Additional model preperation


We take *ufo_model_data* which was the main dataframe, and we join the "places" dataframe to it.
```{r}
ufo_final_model_data <- ufo_model_data %>%
  left_join(places_model_data, by = c("city", "state", "country_code"))

ufo_final_model_data
```


We have selected only several columns that we determined to be the most necessary in our predictions: shape, duration (in seconds), part of day, population of the city it took place and the geographical data.
```{r}
ufo_final_model_data <- ufo_final_model_data %>%
  select(shape, duration_seconds, day_part, population, elevation_m, latitude, longitude) %>%
  filter(!is.na(shape)) %>%
  mutate(shape = factor(shape))

ufo_final_model_data
```


We wanted our model to predict the type of shape a UFO sighting could have. This dataset had a lot of different shapes, so we decided to only use the 4 most common shapes
```{r}
top_shapes <- ufo_final_model_data %>%
  count(shape, sort = TRUE) %>%
  slice(1:4) %>%
  pull(shape)

top_shapes
```

Filtering the dataframe based on the 4 most common shapes
```{r}
ufo_final_model_data <- ufo_final_model_data %>%
  filter(shape %in% top_shapes) %>%
  mutate(shape = factor(shape))

ufo_final_model_data
```


Five fold cross validation set-up
```{r}
set.seed(123)
split <- initial_split(ufo_final_model_data, strata = day_part)
train_data <- training(split)
test_data <- testing(split)

```


Recipe
```{r}
rec <- recipe(shape ~ duration_seconds + day_part + population + elevation_m + latitude + longitude,
              data = ufo_final_model_data) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors()) %>%
  step_normalize(all_numeric_predictors())

rec
```

#Random Forest

Our second model type is a random forest. We tried it as it also fit our case and would be (in theory) more advanced than multinomal regression.

Model
```{r}
rf_model <- rand_forest(mode = "classification", trees = 500) %>%
  set_engine("ranger")

rf_model
```

Workflow
```{r}
wf_rf <- workflow() %>%
  add_recipe(rec) %>%
  add_model(rf_model)
```


Model fitting
```{r}
rf_fit <- fit(wf_rf, data = train_data)
```


Predictions
```{r}
rf_preds <- predict(rf_fit, test_data) %>%
  bind_cols(test_data) %>%
  mutate(
    shape = factor(shape),
    .pred_class = factor(.pred_class, levels = levels(shape))
  )
```

<<<<<<< HEAD

Trained model metrics. Generally pretty poor results
=======
>>>>>>> cb33d2047023d675e060afd2b401560b5c0ebe59
```{r}
metrics <- metric_set(accuracy, precision, recall, f_meas)

metrics(rf_preds, truth = shape, estimate = .pred_class, estimator = "macro_weighted")
```

<<<<<<< HEAD


Confusion matrix - it clearly shows that the model does not want to predict anything other than *circle* and *light*. This time however there is definetly an improvement.
=======
>>>>>>> cb33d2047023d675e060afd2b401560b5c0ebe59
```{r}
conf_mat(rf_preds, truth = shape, estimate = .pred_class)
```

<<<<<<< HEAD


=======
## Step 1: Data Splitting

```{r}
final_data_split <- initial_split(final_data, prop = 3/4, strata = shape)

data_train <- training(final_data_split)
data_test <- testing(final_data_split)

cat("Training data:", nrow(data_train), "observations\n")
cat("Testing data:", nrow(data_test), "observations\n")

final_data <- vfold_cv(data_train, v = 10, strata = shape)
```

## Step 2: Preprocessing

```{r}
data_recipe <- recipe(
    shape ~ .,
    data = data_train) %>% 
  step_normalize(all_numeric_predictors())
```

```{r}
summary(data_recipe)
```

## Step 3: Model Specification

```{r}
rf_spec <- rand_forest(mode = "classification") %>% 
  set_engine("ranger") 

rf_spec
```

## Step 4: Bundling into a Workflow

```{r}
data_wf <- workflow() %>% 
  add_recipe(data_recipe) %>% 
  add_model(rf_spec)

data_wf
```

## Step 5: Training the Model

```{r fit-model}
data_fit <- fit(data_wf, data = data_train)

data_fit
```

```{r}
predictions_class <- predict(data_fit, new_data = data_test, type = "class")
predictions_class
```

```{r}
predictions_prob <- predict(data_fit, new_data = data_test, type = "prob")
predictions_prob
```

```{r}
results_df <- bind_cols(
  predictions_class,
  predictions_prob,
  data_test %>% select(shape)
) %>%
  mutate(shape = as.factor(shape))

results_df
```

```{r}
classification_metrics <- metric_set(
  accuracy, 
  roc_auc,
  precision,
  recall)
classification_metrics 
```

```{r}

eval_results <- results_df %>%
  classification_metrics(
    truth    = shape,
    estimate = .pred_class,
    dplyr::starts_with(".pred_") & !dplyr::matches("^\\.pred_class$"),
    estimator = "macro"
  )

print(eval_results)
```

## Step 6: Calculate and plot confusion matrix

```{r}
cm <- conf_mat(results_df, truth = shape, estimate = .pred_class)

df_cm <- as.data.frame.table(cm$table)
names(df_cm) <- c("Predicted", "True", "Count")

df_cm$Predicted <- factor(df_cm$Predicted, levels = rev(levels(df_cm$Predicted)))

df_cm <- df_cm %>%
  mutate(text_col = ifelse(Count > max(Count) * 0.5, "white", "black"))

p <- ggplot(df_cm, aes(x = True, y = Predicted, fill = Count)) +
  geom_tile(color = "grey80", size = 0.2) +
  geom_text(aes(label = Count, color = text_col),
            size = 3, show.legend = FALSE) +
  scale_fill_gradient(low = "white", high = "steelblue", name = "Liczba") +
  scale_color_identity() +
  coord_equal() +
  labs(
    title = "Confusion Matrix: True vs. Predicted Shape",
    x     = "True Shape",
    y     = "Predicted Shape"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.background  = element_rect(fill = "white", color = NA),
    plot.background   = element_rect(fill = "white", color = NA),
    legend.background = element_rect(fill = "white", color = NA),
    axis.text.x       = element_text(angle = 45, hjust = 1),
    panel.grid        = element_blank(),
    plot.title        = element_text(hjust = 0.5, face = "bold")
  )

ggsave("conf_mat_rf.png", plot = p,
       width = 12, height = 10, dpi = 300,
       bg = "white")

```
>>>>>>> cb33d2047023d675e060afd2b401560b5c0ebe59
